A Large-scale Study of Polymorphism in R

### 2019-03-31 notebooks/paper_L123_results_section.Rmd
some more results prose, metaprogramming discussion might have to change with new run results

# There's at least one transformation that we're always gonna want to apply, namly the vector/scalar
# collapse where both are present in a signature.
los <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
# Figure out which ones are matrices. Remove if the analysis picks this up (and we rerun).
los <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
# remove "errors"
los <- lapply(los, function(lofun) lapply(lofun, remove_err_df))
```

@@

>>>>

# There's at least one transformation that we're always gonna want to apply, namly the vector/scalar
# collapse where both are present in a signature.
los <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
# Figure out which ones are matrices. Remove if the analysis picks this up (and we rerun).
los <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
# remove "errors"
# TODO: at a later stage, we should show results about the errors
los <- lapply(los, function(lofun) lapply(lofun, remove_err_df))
```

@@

3. Extras. What about errors? How often do we have metaprogramming.
```{r}
meta_los <- lapply(los, function(lof) lapply(lof, annihilate_errors_df))
```

### 2019-03-31 paper/main.tex
some more results prose, metaprogramming discussion might have to change with new run results

\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  framexleftmargin=-1mm,
  framexrightmargin=-1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
 % numbers=left
}

@@

In this section, we detail our methodology for collecting data.  Our aim is
to observe arguments and return values of function calls, and from these
generalize possible type signatures for the called functions.  We base our
infrastructure on an open source tool called \genthat whose purpose is to
generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
by synthesizing unit tests from recorded function argument and return
values, comparing the test against existing ones to avoid generating tests
which do not increase code coverage.  To suit our purposes, we change the
existing tool in two main ways: we record \emph{shapes} rather than values,
and ignore the code coverage optimization phase.  Both changes are
beneficial for scalability, allowing us to trace far more calls.

@@

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L                     
[1] 3
> x <- c(1,2,3,4)            
> attr(x,"dim") <- c(2,2)    
> x + c(1,2)                 
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} & 
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+` &\tt D[1], I[1] -> D[1] \\
\tt c& D[1], D[1], D[1], D[1] -> D[4]\\
\tt `<-` &\tt S, D[3] -> D[3]\\
\tt c & \tt D[1], D[1] -> D[2]\\
\tt `attr<-1`&\tt  S, C, D[2] -> \\
\tt c &\tt D[1], D[1] -> D[2]\\
\tt `+` &\tt D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}.  The shapes we would expect to record are as follows:
In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
\k{character}. For vectorized types we record their length and for all types
we record attributes with some of their values. Looking at the signatures
observed for addition, it is clear that the function is polymorphic as it
starts with the addition of two scalar numbers of different types, and then
adds a matrix to a vector returning a matrix.

@@

In R all arguments are passed as promises, and unused arguments will be
unevaluated.  Our tracer does still collect information on these arguments,
as whether or not they were used doesn't change the fact that they were in
fact passed.  In some cases, arguments aren't even intended to be evaluated.
For example, consider the {\tt magrittr} package, which implements the
Tidyverse's pipe functions.  Refer to Figure~\ref{fig:magpipeex} for a
translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by <- function(x, y) { x * y }

4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}

Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)},
but that's erroneous code (you can't call {\tt mul\_by} without supplying
two arguments).  In this situation, our tracer will still register the
entire call, but flag the {\tt rhs} argument to indicate to later stages of
the analysis that there was an issue evaluating that argument.

>>>>

\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  framexleftmargin=-1mm,
  framexrightmargin=-1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  numbers=left
}

@@

In this section, we will detail our methodology for collecting data.  
Basically, our aim is to observe arguments and return values of function calls, and from these generalize possible type signatures for the called functions. 
To this end, we build on an open source tool called \genthat whose purpose is to generate unit tests for R libraries~\cite{issta18}.  
\genthat achieves this by synthesizing unit tests from recorded function argument and return values, comparing the test against existing ones to avoid generating tests which do not increase code coverage.
To suit our purposes, we change the existing tool in two main ways: 
We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run \genthat's code coverage minimization phase. 
Both changes were beneficial for scalability, allowing us to trace far more calls than was reported in the \genthat work~\cite{issta18}.

\subsection{Example}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 1 + 2L                        # add double and integer
[1] 3
> x <- c(1,2,3,4)               # save double vector
> attr(x,"dim") <- c(2,2)       # assign to "dim" attribute
> x + c(1,2)                    # add double 3 to each matrix element matrix
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}\caption{Example script.}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}.  Some functions syntactic support \AT{?}, addition is infix, and the
assignment to an attribute is written mixfix. The shapes we would expect to
record are as follows:

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
`+`: D[1], I[1] -> D[1] # (args: double size 1, integer size 1, return: double size 1)
c: D[1], D[1], D[1], D[1] -> D[4]
`<-`: S, D[3] -> D[3]
c: D[1], D[1] -> D[2]
`attr<-1`:  S, C, D[2] -> 
c: D[1], D[1] -> D[2]
`+`: D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
\end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}

In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
\k{character}. For vectorized types we record their length and for all types
we record attributes with some of their values. Looking at the signatures
observed for addition, it is clear that the function is polymorphic as it
starts with the addition of two scalar numbers of different types, and then
adds a matrix to a vector returning a matrix.

@@

In R all arguments are passed as promises, and unused arguments will be unevaluated.
Our tracer does still collect information on these arguments, as whether or not they were used doesn't change the fact that they were in fact passed. 
In some cases, arguments aren't even intended to be evaluated.
For example, consider the {\tt magrittr} package, which implements the Tidyverse's pipe functions.
Refer to Figure~\ref{fig:magpipeex} for a translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by <- function(x, y) { x * y }

4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}
Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)}, but that's erroneous code (you can't call {\tt mul\_by} without supplying two arguments).
In this situation, our tracer will still register the entire call, but flag the {\tt rhs} argument to indicate to later stages of the analysis that there was an issue evaluating that argument.
